\documentclass[a4paper,12pt,fleqn]{article}

\usepackage{amsmath,amsthm,amssymb}
\usepackage[margin=1in]{geometry}
\usepackage{xcolor}
\usepackage{array}
\usepackage{comment}
%\usepackage{booktabs}
\usepackage[bookmarksnumbered=true,hypertexnames=false]{hyperref}
%\usepackage{algorithm, algpseudocode}
\usepackage[capitalize,sort]{cleveref}

%\def\colorscheme{dark}
\input{colorscheme.tex}
\hypersetup{colorlinks,linkcolor=textRed,citecolor=textRed,urlcolor=textBlue}
\input{macros.tex}

\author{Eklavya Sharma}
\date{\empty}

\title{Linear Programming Counterexamples}

\begin{document}

\maketitle
\setlength{\parskip}{0.2em}

\section{Polyhedra}

\begin{example}[Square Pyramid]
\label{ex:pyramid-1}
Let
\begin{align*}
P &\defeq \{(x, y, z): \max(|x|, |y|) \le z, z \le 1\}
\\ &= \{(x, y, z): x \le z, -x \le z, y \le z, -y \le z, z \le 1\}.
\end{align*}
Then $P$ is an (inverted) square pyramid.
The base of the pyramid is $[-1, 1]^2 \times \{1\}$.
Its vertices are $(0, 0, 0)$, $(-1, -1, 1)$, $(-1, 1, 1)$, $(1, -1, 1)$, $(1, 1, 1)$.
\end{example}

\begin{example}[Square Pyramid]
\label{ex:pyramid-2}
Let
\[ P \defeq \{(x, y, z): x \le z, y \le z, x \ge 0, y \ge 0, z \ge 0\}. \]
Then $P$ is an (inverted) square pyramid.
The base of the pyramid is $[0, 1]^2 \times \{1\}$.
Its vertices are $(0, 0, 0)$, $(0, 0, 1)$, $(0, 1, 1)$, $(1, 0, 1)$, $(1, 1, 1)$.
\end{example}

\begin{example}[Introducing degeneracy by changing RHS to 0]
Let $P \defeq \{x: (a_i^Tx = b_i \forall i \in E) \textrm{ and } (a_i^Tx \le b_i \forall i \in E)\}$.
Let $D \defeq \{x: (a_i^Tx = 0 \forall i \in E) \textrm{ and } (a_i^Tx \le 0 \forall i \in E)\}$.
Then $D$ is the set of directions of $P$. If $P$ is bounded, then $D = \{0\}$.
Now all bases of $D$ correspond to the same point.
If $P$ has multiple bases, then $0$ is a degenerate point of $D$.
Furthermore, the simplex method can be made to run on $P$ and $D$ with the same pivots.
\end{example}

\begin{example}[Non-extreme point with $n$ active constraints]
Let $P \defeq \{(x, y): x + y \ge 1, x + y \le 1, x \ge 0, y \ge 0\}$.
No constraint is redundant.
$(1,0)$ and $(0,1)$ are degenerate BFSes of $P$.
Their midpoint, $(1/2, 1/2)$, is not an extreme point, but has 2 active constraints
(which are linearly dependent).
\end{example}

\subsection{Degeneracy vs Redundancy}

\begin{example}[Redundancy $\nRightarrow$ Degeneracy]
Let $P \defeq \{(x, y): 0 \le x \le 1, 0 \le y \le 1\}$.
Then every extreme-point is non-degenerate.
Adding the constraint $y \le 2$ doesn't add degeneracy but adds redundancy.
\end{example}

\begin{example}[Degeneracy $\nRightarrow$ Redundancy]
Let $P \defeq \{(x, y): y \le x, y \le -x, y \ge 0\}$ (so $P = \{(0, 0)\}$).
Then $(0, 0)$ is a degenerate extreme point, but no constraint is redundant.
\end{example}

\begin{example}[Degeneracy $\nRightarrow$ Redundancy]
Let $P$ be a square pyramid (c.f.~\cref{ex:pyramid-1}).
Then there is a degenerate extreme point but no constraint is redundant.
\end{example}

\section{Simplex Method}

\begin{example}
Let $b \in \mathbb{R}_{\ge 0}$ and $a \in \mathbb{R}^n$, where $0 < a_1 < a_2 < \ldots < a_n$.
Consider the LP \[ \max_{x \ge 0}\quad a^Tx \quad\textrm{where}\quad \sum_{i=1}^n x_i \le b. \]
Clearly, the optimal solution is $[0, 0, \ldots, 0, b]$.
In standard form, the LP becomes
\[ \min_{x \ge 0}\quad \sum_{i=1}^n (-a_i)x_i \quad\textrm{where}\quad \sum_{i=1}^{n+1} x_i = b. \]
Suppose our initial basis is $\{x_{n+1}\}$. For ease of notation, let $x_0 \defeq x_{n+1}$.

If we run the simplex method with Bland's rule (variable of lowest index enters basis),
then there will be $n$ iterations, where in the $i\Th$ iteration,
$x_i$ enters the basis and $x_{i-1}$ leaves the basis.
Hence, we visit each of the $n+1$ bases.
If $b > 0$, each basis corresponds to a unique BFS.
If $b = 0$, all bases correspond to the same BFS $\mathbf{0}$.

If we run the simplex method with Dantzig's rule (variable of most negative reduced cost enters basis),
then there will be just 1 iteration where $x_n$ enters and $x_{n+1}$ leaves.
\end{example}

\begin{example}
For the following LP, where $b \in \mathbb{R}_{\ge 0}$, the optimal solution is $(b, 0)$.
\[ \max_{x \ge 0, y \ge 0}\quad 2x+3y \quad\textrm{where}\quad x + 2y \le b. \]
Bland's rule takes 1 iteration but Dantzig's rule takes 2 iterations.
There are 3 bases. When $b > 0$, each base corresponds to a different BFS.
When $b = 0$, all bases correspond to the same BFS $\mathbf{0}$.
\end{example}

\begin{example}[Unique basis and degeneracy]
Let $P \defeq \{(x, y, z): x + y = 2b, x - y = 0, x \ge 0, y \ge 0, z \ge 0\}$,
be a standard form polyhedron. Then $\{x, y\}$ is the only basis,
and the corresponding solution is $(b,b,0)$.
The solution is degenerate iff $b = 0$.
\end{example}

%\addMyBib{}

\end{document}
