\input{header.tex}

\title{Bounds on Sorting}

\begin{document}

\maketitle

\begin{abstract}
This document analyzes lower and upper bounds on the worst-case number of comparisons required for sorting
an array of $n$ elements. This is done for both sorting in general and for specific algorithms.
\end{abstract}

\initFromContents{}

\section{General lower bound}

By the decision tree model of computing, we get a lower bound of $\ceil{\lg(n!)}$
on the number of comparisons in the worst case.

By Stirling's approximation, we get
\[ \lg(n!) = n\lg n - (\lg e)n + \frac{1}{2}\lg n + \lg\sqrt{2\pi}
+ (\lg e)\left[ \frac{1}{12n+1}, \frac{1}{12n} \right] \]
($\lg e \approx 1.4427$ and $\lg\sqrt{2\pi} \approx 1.3257$)

\section{Specific algorithms}

\subsection{Insertion sort}

In the worst case, insertion sort performs $\frac{n(n-1)}{2}$ comparisons.

\subsection{Insertion sort with binary search}

Binary searching an array of size $n$ takes $\floor{\lg n} + 1$ comparisons.
(Solve the recurrence $f(1) = 1 \wedge f(n) = f\left(\floor{\frac{n}{2}}\right) + 1$)

Therefore, number of comparisons is
\[ (n-1) + \sum_{i=1}^{n-1} \floor{\lg i} \le (n-1) + \floor{\lg((n-1)!)} \]

\subsection{Merge sort}

Merging 2 sorted arrays of size $m$ and $n$ can be done in at most $m + n - 1$ comparisons.

In the worst case, merge sort performs $f(n)$ comparisons, where
$f(0) = f(1) = 0$ and
$f(n) = f\left(\floor{\frac{n}{2}}\right) + f\left(\ceil{\frac{n}{2}}\right) + (n-1)$.

The solution to this recurrence is \cite{eku-notes-recurrences}
\[ f(n)
= n(\floor{\lg n} + 1) - 2^{\floor{\lg n} + 1} + 1
\in n\floor{\lg n} - [0, n-1] \]
This is $O(n)$ higher than the decision-tree lower bound.

\subsection{Heapsort}

With a binary heap, total number of comparisons for heapsort
\[ \le 2(n-1 + \floor{\lg((n-1)!)}) \le 2n\lg n - 2(\lg e - 1)n - \lg n + \lg\pi - \frac{5}{6} \]
See \cite{eku-notes-heaps} for the algorithm and analysis.

\subsection{Randomized quicksort}

Partitioning an array of size $n$ about a pivot can be done in $n-1$ comparisons.

Let $f(n)$ be the expected number of comparisons required for randomized quicksort.
Therefore, $f(0) = f(1) = 0$ and
\[ f(n) = (n-1) + \frac{1}{n} \sum_{i=1}^n (f(i-1) + f(n-i)) \]

\begin{align*}
& f(n) = (n-1) + \frac{1}{n} \sum_{i=1}^n (f(i-1) + f(n-i))
\\ &\Rightarrow nf(n) = n(n-1) + 2 \sum_{i=0}^{n-1} f(i-1)
\\ &\Rightarrow nf(n) - (n-1)f(n-1) = 2(n-1) + 2f(n-1) \tag{subtract equations for $n$ and $n-1$}
\\ &\Rightarrow \frac{f(n)}{n+1} - \frac{f(n-1)}{n}
= \frac{2(n-1)}{n(n+1)} = \frac{4}{n+1} - \frac{2}{n}
\\ &\Rightarrow \frac{f(n)}{n+1} - f(0)
= \sum_{i=1}^n \left( \frac{4}{i+1} - \frac{2}{i} \right)
= 2H(n+1) + \frac{2}{n+1} - 4 \tag{$H(n) = \sum_{i=1}^n \frac{1}{i}$}
\\ &\Rightarrow f(n) = 2((n+1)H(n) - 2n)
\end{align*}

Using the integration bound for the sum of a decreasing function:
\[ \sum_{i=a}^b f(i) \in \left( \int_a^b f(x)dx \right) + [f(b), f(a)] \]
we get $H(n) \in \ln n + \left[ \frac{1}{n}, 1 \right]$.

Therefore,
\[ f(n) = 2((n+1)H(n) - 2n) \le 2n\ln n - 2n + 2H(n) \]
\[ \le \left(\frac{2}{\lg e}\right)n\lg n - 2n + 2H(n) \]
Since, $\frac{2}{\lg e} \approx 1.3863$, randomized quicksort takes
approximately 1.3863 times the number of comparisons by the decision-tree lower bound.

\addMyBib{}

\end{document}
