\documentclass[a4paper, 12pt, fleqn]{article}

\usepackage{amsmath,amsthm,amssymb}
\usepackage[margin=1in]{geometry}
\usepackage{xcolor}
\usepackage{array}
\usepackage{comment}
\usepackage{booktabs}
\usepackage[bookmarksnumbered=true,hypertexnames=false]{hyperref}
\usepackage{algorithm, algpseudocode}
\usepackage[capitalize,sort]{cleveref}

%\def\colorscheme{dark}
\input{colorscheme.tex}
\hypersetup{colorlinks,linkcolor=textRed,citecolor=textRed,urlcolor=textBlue}
\input{macros.tex}

\DeclareMathOperator{\idFunc}{id}

\author{\empty}
\date{\empty}

\title{Chapter 1: Preliminaries}

\begin{document}

\maketitle
\setlength{\parskip}{0.2em}

\section{Sets}

\begin{definition}[Set basics]
\leavevmode
\begin{enumerate}
\item $A \subseteq B \defiff (\forall x \in A, x \in B)$.
\item $A = B \defiff (A \subseteq B \land B \subseteq A)$.
\item $A \subset B \defiff (A \subseteq B \land B \not\subseteq A)$.
\item $A \cup B \defeq \{x: x \in A \textrm{ or } x \in B\}$.
\item $A \cap B \defeq \{x \in A: x \in B\}$.
\item $A \setminus B \defeq \{x \in A: x \not\in B\}$.
\item $\displaystyle \bigcup_{i \in I} A_i \defeq \{x: \exists i \in I \textrm{ such that } x \in A_i\}$.
\item $\displaystyle \bigcap_{i \in I} A_i \defeq \{x: \forall i \in I, x \in A_i\}$.
\item $A \times B \defeq \{(x, y): x \in A, y \in B\}$.
\item $\displaystyle \prod_{i=1}^n A_i \defeq \{(x_1, x_2, \ldots, x_n): x_i \in A_i \textrm{ for all } i\}$.
\end{enumerate}
\end{definition}

\begin{theorem}
\begin{enumerate}
\item $A \subseteq B \iff A \cap B = A \iff A \cup B = B$.
\item $A \setminus (B \cup C) = (A \setminus B) \cap (A \setminus C)$.
\item $A \setminus (B \cap C) = (A \setminus B) \cup (A \setminus C)$.
\item $A \cup (B \cup C) = (A \cap B) \cup (A \cap C)$.
\item $A \cap (B \cap C) = (A \cup B) \cap (A \cup C)$.
\end{enumerate}
\end{theorem}

\section{Relations and Functions}

\begin{definition}[Relation and function]
A relation $R$ between $A$ and $B$ is a subset of $A \times B$.
A function $f: A \to B$ is a relation between $A$ and $B$ such that
\[ (a, b_1) \in f \textrm{ and } (a, b_2) \in f \implies b_1 = b_2. \]
$D(f) \defeq A$ (called \emph{domain} of $f$),
and $R(f) \defeq B$ (called \emph{range} of $f$).
\end{definition}

\begin{lemma}
Let $f: A \to B$ and $g: A \to B$.
Then $f = g \iff (\forall x \in A, f(x) = g(x))$.
\end{lemma}

\begin{definition}[Image and reverse image]
Let $f: A \to B$ be a function.
\begin{enumerate}
\item For $X \subseteq A$, $f(X) \defeq \{f(x): x \in X\}$ is called the \emph{image} of $X$ under $f$.
    \\ Equivalently, $y \in f(X) \iff (\exists x \in X, f(x) = y)$.
\item For $Y \subseteq B$, $f^{-1}(Y) = \{x: f(x) \in Y\}$ is called the \emph{inverse image}
    of $Y$ under $f$. Equivalently, $x \in f^{-1}(Y) \iff f(x) \in Y$.
\end{enumerate}
\end{definition}

\begin{lemma}
Let $f: A \to B$. Let $X_1, X_2 \subseteq A$ and $Y_1, Y_2 \subseteq B$.
\begin{enumerate}
\item $f(X_1 \cup X_2) = f(X_1) \cup f(X_2)$.
\item $f(X_1 \cap X_2) \subseteq f(X_1) \cap f(X_2)$.
\item $f^{-1}(Y_1 \cup Y_2) = f^{-1}(Y_1) \cup f^{-1}(Y_2)$.
\item $f^{-1}(Y_1 \cap Y_2) = f^{-1}(Y_1) \cap f^{-1}(Y_2)$.
\end{enumerate}
\end{lemma}

\begin{definition}[Composition]
\label{defn:func-compose}
For functions $f: A \to B$ and $g: B \to C$,
$g \circ f: A \to C$ is the defined as $(g \circ f)(x) = g(f(x))$.
\end{definition}

\begin{definition}[Injection and surjection]
Let $f: A \to B$.
\begin{enumerate}
\item $f$ is injective (aka one-to-one)
    $\defiff \forall x_1 \in A, \forall x_2 \in A, (f(x_1) = f(x_2) \implies x_1 = x_2)$.
\item $f$ is surjective (aka onto)
    $\defiff \forall y \in B, \exists x \in A, f(x) = y$.
\end{enumerate}
\end{definition}

\begin{lemma}[Composition]
\label{thm:inj-surj-comp}
Let $f: A \to B$ and $g: B \to C$.
\begin{enumerate}
\item If $f$ and $g$ are injective, then $g \circ f$ is injective.
\item If $g \circ f$ is injective, then $f$ is injective.
\item If $f$ and $g$ are surjective, then $g \circ f$ is surjective.
\item If $g \circ f$ is surjective, then $g$ is surjective.
\end{enumerate}
\end{lemma}

\begin{definition}[Identity]
The identity function $\idFunc_A: A \to A$ is given by $\idFunc_A(x) = x$ for all $x \in A$.
\end{definition}

\begin{definition}[Bijection]
A function $f: A \to B$ is a bijection iff (the following are equivalent):
\begin{enumerate}
\item \label{item:bij:inj-surj}$f$ is injective and surjective.
\item \label{item:bij:inv}$\exists g: B \to A$ such that $g \circ f = \idFunc_A$ and $f \circ g = \idFunc_B$.
    (Then $g$ is called the \emph{inverse} of $f$, and is denoted by $f^{-1}$.)
\end{enumerate}
\end{definition}
\begin{proof}[Proof sketch of equivalence]
If $f$ is injective and surjective, for each $y \in B$, there is a unique $x \in A$
such that $f(x) = y$. Define $g(y) = x$ and show condition \ref{item:bij:inv}.
To show that condition \ref{item:bij:inv} implies condition \ref{item:bij:inj-surj},
use \cref{thm:inj-surj-comp}.
\end{proof}

\begin{definition}[Restriction]
Let $f: A \to B$ be a function. Let $X \subseteq A$. Then $f|X$ is a function from $X$ to $B$
such that $(f|X)(x) = f(x)$ for all $x \in X$.
\end{definition}

\begin{lemma}
There is an injection from $A$ to $B$ iff there is a surjection from $B$ to $A$.
\end{lemma}
\begin{proof}
$\implies$:
\\ Let $f: A \to B$ be injective.
Then for any $y \in B$, $f^{-1}(y)$ contains 0 or 1 elements.
Let $a$ be an arbitrary element of $A$. Define $g: B \to A$ as:
\[ g(y) = \begin{cases}x & \textrm{ if } f^{-1}(y) = \{x\}
\\ a & \textrm{ if } f^{-1}(y) = \emptyset \end{cases}. \]
Pick any $x \in A$. Let $y = f(x)$. Then $x \in f^{-1}(y)$. Hence, $g(y) = x$.
Hence, $g$ is surjective.

$\impliedby$:
\\ Let $g: B \to A$ be surjective.
Then for any $x \in A$, we have $g^{-1}(x) \neq \emptyset$.
By the axiom of choice, there is a function $f: A \to B$ such that $f(x) \in g^{-1}(x)$.
For any distinct $x_1, x_2 \in A$, we have $g^{-1}(x_1) \cap g^{-1}(x_2) = \emptyset$.
Hence, $f$ is injective.
\end{proof}

\section{Set Cardinality}

\begin{definition}
A set $S$ is \emph{countable} iff there exists a surjection $f: \mathbb{N} \to S$.
A set $S$ is \emph{denumerable} iff it is countable and infinite.
\end{definition}

\begin{lemma}
A set $S$ is \emph{countable} iff there exists a surjection $f: T \to S$, where $T$ is countable.
\end{lemma}

\begin{lemma}
There is a bijection from $\mathbb{N}$ to $\mathbb{N} \times \mathbb{N}$.
(Hence, $\mathbb{N} \times \mathbb{N}$ is countable.)
\end{lemma}
\begin{proof}[Proof sketch]
The bijection $f: \mathbb{N} \to \mathbb{N} \times \mathbb{N}$ is defined as
\begin{algorithmic}[1]
\State Set $j = 1$
\For{$s \in \mathbb{N}$}
    \For{$i$ from $1$ to $s-1$}
        \State Set $f(j) = (i, s-i)$
        \State $j$ \texttt{+=} 1
    \EndFor
\EndFor
\end{algorithmic}
\end{proof}

\begin{theorem}
\label{thm:prod-countable}
Let $I$ be a finite set and for each $i \in I$, let $A_i$ be a countable set.
Then $\prod_{i \in I} A_i$ is countable.
\end{theorem}
\begin{proof}[Proof sketch]
When $|I| = 2$, we can show a surjection from $\mathbb{N} \times \mathbb{N}$ to $A_1 \times A_2$.
For larger $|I|$, use induction on $|I|$.
\end{proof}

\begin{theorem}
\label{thm:union-countable}
Let $I$ be a countable set and for each $i \in I$, let $A_i$ be a countable set.
Then $\bigcup_{i \in I} A_i$ is countable.
\end{theorem}
\begin{proof}[Proof sketch]
Since $I$ is countable, there exists a surjection $f: \mathbb{N} \to I$.
Since $A_i$ is countable, there exists a surjection $g_i: \mathbb{N} \to A_i$.
Then $h(i, j) = g_i(j)$ is a surjection from $\mathbb{N} \times \mathbb{N}$ to $\bigcup_{i \in I} A_i$.
\end{proof}

\begin{theorem}
\label{thm:finite-subsets-countable}
Let $A$ be a countable set.
Let $B$ be the set of all finite subsets of $A$. Then $B$ is countable.
\end{theorem}
\begin{proof}[Proof sketch]
Express $B$ as union of $k$-sized sets and use \cref{thm:union-countable}.
For each $k$, a subset is finite by \cref{thm:prod-countable}.
\end{proof}

\begin{theorem}[Cantor's theorem]
Let $P$ be the power set of $A$.
Then there is no surjection from $A$ to $P$.
\end{theorem}
\begin{proof}[Proof sketch]
Assume there is a surjection $f: A \to P$.
Let $D \defeq \{a \in A: a \not\in f(a)\}$.
Since $D \in P$ and $f$ is a surjection, we have $D = f(a_0)$ for some $a_0 \in A$.
For each case $a_0 \in D$ and $a_0 \not\in D$, show a contradiction.
\end{proof}

\end{document}
