\documentclass[a4paper, 12pt, fleqn]{article}

\usepackage{amsmath,amsthm,amssymb}
\usepackage[margin=1in]{geometry}
\usepackage{xcolor}
\usepackage{array}
\usepackage{comment}
\usepackage[bookmarksnumbered=true,hypertexnames=true]{hyperref}
\usepackage{algorithm, algpseudocode}
\usepackage[capitalize,sort]{cleveref}

%\def\colorscheme{dark}
\input{colorscheme.tex}
\hypersetup{colorlinks,linkcolor=textRed,citecolor=textRed,urlcolor=textBlue}
\input{macros.tex}

\author{\empty}
\date{\empty}

\title{Chapter 2: Real numbers}

\begin{document}

\maketitle
\setlength{\parskip}{0.2em}

\section{Groups}

\begin{definition}[Group]
Let $G$ be a non-empty set and $\circ: G \times G \to G$ be a binary operator.
Then $(G, \circ)$ is a group iff all of the following hold:
\begin{enumerate}
\item Associativity: $a \circ (b \circ c) = (a \circ b) \circ c$ for all $a, b, c \in G$.
\item Identity exists: $\exists e \in G$ such that $\forall a \in G$, $e \circ a = a \circ e = a$.
    Such an $e$ is called an \emph{identity} of $(G, \circ)$.
    We can prove that the identity is unique.
\item Inverses exist: Let $e$ be an identity of $(G, \circ)$. Then $\forall a \in G$,
    $(\exists \ell \in G, \ell \circ a = e)$ and $(\exists r \in G, a \circ r = e)$.
    $\ell$ is called a \emph{left inverse} of $a$. $r$ is called a \emph{right inverse} of $a$.
\end{enumerate}
$(G, \circ)$ is called \emph{symmetric}, \emph{commutative}, or \emph{abelian} iff
$\forall a \in G$, $\forall b \in G$, $a \circ b = b \circ a$.
\end{definition}

\begin{lemma}
In a group $(G, \circ)$, the identity is unique and each element has a unique inverse.
\end{lemma}
\begin{proof}
Let $e_1$ and $e_2$ be identities of $(G, \circ)$.
Then $e_1 \circ e_2 = e_1$, since $e_2$ is an identity, and $e_2 \circ e_1 = e_2$, since $e_1$ is an identity.
Hence, $e_1 = e_2$.

Let $\ell$ be a left inverse and $r$ be a right inverse of $a \in G$. Then
\[ \ell = \ell \circ e = \ell \circ (a \circ r) = (\ell \circ a) \circ r = e \circ r = r. \]
Hence, every left inverse equals every right inverse. Hence, they are all equal.
\end{proof}

\begin{definition}[Standard operators]
If we use $+$ as a group operator, we denote identity as 0 and inverse of $g$ as $-g$.
If we use $\times$ as a group operator, we denote identity as 1 and inverse of $g$ as $g^{-1}$.
$a - b \defeq a + (-b)$. $a / b \defeq ab^{-1}$.
\end{definition}

\begin{definition}
Let $(G, \times)$ be a group. Then for any $n \in \mathbb{Z}$ and any $g \in G$, define
\[ g^n = \begin{cases}
g \times g \times \ldots \times g \; (n \textrm{ times}) & \textrm{if } n > 0
\\ 1 & \textrm{if } n = 0
\\ g^{-1} \times g^{-1} \times \ldots \times g^{-1} \; (-n \textrm{ times}) & \textrm{if } n < 0
\end{cases}. \]
\end{definition}

\begin{lemma}[Basic properties]
Let $(G, \cdot)$ be a group. Let $a, b \in G$ and $m, n \in \mathbb{Z}$.
\begin{enumerate}
\item $(ab)^{-1} = b^{-1}a^{-1}$.
\item $(a^{-1})^{-1} = a$.
\item $a^ma^n = a^{m+n}$.
\item $(a^m)^n = a^{mn}$.
\item If $G$ is symmetric, $(ab)^n = a^nb^n$.
\end{enumerate}
\end{lemma}

\section{Fields}

\begin{definition}[Field]
$(F, +, \times)$ is a field iff it satisfies all of the following:
\begin{enumerate}
\item $(F, +)$ is a symmetric group. It's identity is denoted as 0.
\item $(F - \{0\}, \times)$ is a symmetric group. It's identity is denoted as 1.
\item Distributivity: $a(b + c) = ab + ac$ and $(a+b)c = ac + bc$.
\end{enumerate}
\end{definition}

\begin{lemma}[Basic properties]
\label{thm:field-basics}
Let $(F, +, \times)$ be a field. Let $a, b \in F$.
\begin{enumerate}
\item $a0 = 0a = 0$.
\item $a(-b) = (-a)b = -(ab)$.
\item $(-a)(-b) = ab$.
\item \label{item:field-basics:zero-prod}$ab = 0 \iff (a = 0 \textrm{ or } b = 0)$.
\item $(-a)^{-1} = -a^{-1}$.
\end{enumerate}
\end{lemma}
\begin{proof}[Proof sketches]
\leavevmode
\begin{enumerate}
\item $a0 = a(0 + 0) = a0 + a0$.
\item $0 = a0 = a(b + (-b)) = ab + a(-b)$.
\item $(-a)(-b) = a(-(-b)) = ab$.
\item Suppose $a \neq 0$. Then $ab = 0 \implies b = a^{-1}0 = 0$.
\item $(-1)(-1) = 1$, so $(-1)^{-1} = -1$. $(-a)^{-1} = ((-1)a)^{-1} = (-1)^{-1}a^{-1} = -a^{-1}$.
\end{enumerate}
\end{proof}

\section{Partial Orders}

\begin{definition}[Partial and total orders]
Let $L$ be a set and let $\le$ be a binary predicate over $L \times L$.
Then $(L, \le)$ is called a \emph{partial order} (aka poset) iff all of the following hold:
\begin{enumerate}
\item Reflexivity: $\forall a \in L$, $a \le a$.
\item Anti-symmetry: $a \le b$ and $b \le a$ $\implies a = b$.
\item Transitivity: $a \le b$ and $b \le c$ $\implies a \le c$.
\end{enumerate}
Additionally, if $\forall a, b \in L$, we have $a \le b$ or $b \le a$,
then $(L, \le)$ is called a \emph{total order}.

$a < b \defiff (a \le b \textrm{ and } a \neq b)$.
$a \ge b \defiff b \le a$. $a > b \defiff b < a$.
\end{definition}

\begin{definition}[Upper and lower bound]
Let $(L, \le)$ be a poset. Let $S \subseteq L$.
\begin{enumerate}
\item $u \in L$ is an \emph{upper bound} for $S$ iff $s \le u$ for all $s \in S$.
    $S$ is called \emph{upper-bounded} iff an upper bound exists for $S$.
\item $u \in L$ is a \emph{least upper bound} or \emph{supremum} for $S$ (denoted $\sup(S)$)
    iff $u$ is an upper bound for $S$ and for every upper bound $v$ of $S$, we have $u \le v$.
\item $u \in L$ is a \emph{lower bound} for $S$ iff $u \le s$ for all $s \in S$.
    $S$ is called \emph{lower-bounded} iff a lower bound exists for $S$.
\item $u \in L$ is a \emph{greatest lower bound} or \emph{infimum} for $S$ (denoted $\inf(S)$)
    iff $u$ is a lower bound for $S$ and for every lower bound $v$ of $S$, we have $v \le u$.
\item $S$ is called \emph{bounded} iff it has a lower bound and an upper bound.
\end{enumerate}
\end{definition}

\begin{lemma}
$\sup(S)$, if it exists, is unique.
$\inf(S)$, if it exists, is unique.
\end{lemma}

\section{Ordered Field}

\begin{definition}[Ordered field]
\label{defn:ord-field}
Let $(F, +, \times)$ be a field.
$(F, +, \times, \le)$ is an ordered field iff all of the following hold:
\begin{enumerate}
\item $(F, \le)$ is a total order.
\item $a \le b \implies (\forall c \in F, a + c \le b + c)$.
\item $a \ge 0$ and $b \ge 0$ $\implies ab \ge 0$.
\end{enumerate}
\end{definition}

\begin{lemma}[Strict inequalities]
Let $(F, +, \times, \le)$ be an ordered field. Then
\begin{enumerate}
\item $a < b$ and $b < c$ $\implies a < c$.
\item $a < b \implies (\forall c \in F, a + c < b + c)$.
\item $a > 0$ and $b > 0$ $\implies ab > 0$.
\end{enumerate}
\end{lemma}
\begin{comment}
\leavevmode
\begin{enumerate}
\item $a = c \implies (a < b \textrm{ and } b < a) \implies \bot$.
\item $a + c = b + c \implies a = b \implies \bot$.
\item $ab = 0 \implies (a = 0 \textrm{ or } b = 0) \implies \bot$ by
    \cref{thm:field-basics}.\ref{item:field-basics:zero-prod}.
\end{enumerate}
\end{comment}

\begin{definition}[Field with positives (non-standard terminology)]
\label{defn:field-with-positives}
Let $(F, +, \times)$ be a field. Let $P \subseteq F$.
$(F, +, \times, P)$ is called a \emph{field with positives} iff
\begin{enumerate}
\item $a, b \in P \implies a + b \in P$.
\item $a, b \in P \implies ab \in P$.
\item $\forall a \in F$, exactly one of these is true:
    $a = 0$, $a \in P$, $-a \in P$.
\end{enumerate}
\end{definition}

The following two results state that either of \cref{defn:ord-field,defn:field-with-positives}
could be used to define the other.

\begin{lemma}
Let $(F, +, \times, P)$ be a field with positives.
Let $a \le b \defiff (b - a \in P \textrm{ or } b = a)$.
Then $(F, +, \times, \le)$ is an ordered field.
\end{lemma}

\begin{lemma}
Let $(F, +, \times, \le)$ be an ordered field. Let $P \defeq \{x \in F: x > 0\}$.
Then $(F, +, \times, P)$ is a field with positives.
\end{lemma}

\begin{lemma}
Let $(F, +, \times, \le)$ be an ordered field.
\begin{enumerate}
\item $a_1 \le b_1$ and $a_2 \le b_2 \implies a_1 + a_2 \le b_1 + b_2$.
\item $a^2 \ge 0$ and $(a^2 = 0 \iff a = 0)$.
\item $1 > 0$.
\item $ab > 0 \implies (a > 0 \textrm{ and } b > 0) \textrm{ or } (a < 0 \textrm{ and } b < 0)$.
\item $a > 0 \implies a^{-1} > 0$.
\end{enumerate}
\end{lemma}

\begin{lemma}
$(\forall \eps > 0, a \le \eps) \implies a \le 0$.
\end{lemma}

\begin{definition}
$\displaystyle |a| \defeq \begin{cases} a & \textrm{ if } a \ge 0
\\ -a & \textrm{ if } a < 0 \end{cases}$.
\end{definition}

\begin{lemma}
\label{thm:abs-basics}
Let $(F, +, \times, \le)$ be an ordered field.
\begin{enumerate}
\item $|a| \ge 0$ and $(|a| = 0 \iff a = 0)$.
\item $|-a| = |a|$.
\item $|a| \ge a$ and $|a| \ge -a$.
\item \label{item:abs-basics:dbl}Let $c \ge 0$. Then $|a| \le c \iff -c \le a \le c$.
\item $-|a| \le a \le |a|$.
\item $|ab| = |a||b|$.
\item For $a \neq 0$, $|a^{-1}| = |a|^{-1}$.
\end{enumerate}
\end{lemma}

\begin{lemma}[Triangle inequalities]
$||a|-|b|| \le |a + b| \le |a| + |b|$.
\end{lemma}
\begin{proof}
$-|a| \le a \le |a|$ and $-|b| \le b \le |b|$.
Add these to get $-(|a|+|b|) \le a + b \le |a| + |b|$.
By \cref{thm:abs-basics}.\ref{item:abs-basics:dbl}, we get $|a + b| \le |a| + |b|$.

By previous result, $|a| = |(a+b) + (-b)| \le |a+b| + |b|$, so $|a|-|b| \le |a+b|$.
Also, $|b| = |(a+b) + (-a)| \le |a+b| + |a|$, so $-|a+b| \le |a|-|b|$.
Hence, $-|a+b| \le |a|-|b| \le |a+b|$.
By \cref{thm:abs-basics}.\ref{item:abs-basics:dbl}, we get $||a|-|b|| \le |a+b|$.
\end{proof}

\section{Real Numbers}

\begin{definition}
The set of real numbers is an ordered field $(\mathbb{R}, +, \times, \le)$
in which every set with an upper bound has a supremum.
\end{definition}

\end{document}
